Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-10 paperDraft/paperDraft.Rmd
first draft of abstract

To measure bias in word embeddings, @Caliskan2017semanticsBiases proposed the Word Embedding Association Test (WEAT). The idea is that the measure of  biases between two sets of target words, $X$ and $Y$, (we call them protected words) should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we'll call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words. WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. If the person's attitude towards given pair of concept is to be interpreted as neutral, the final value from the formula should be around 0.
Let $f$ be a similarity measure (usually, cosine similarity). The association difference for a term $t$ is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align}
\noindent then, the association difference between $A$ a $B$ is:
\begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align}
\noindent
$s(X,Y,A,B)$ is the statistic used in the signifcance test, and the $p$-value obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$. The effect size is computed by normalizing the difference in means as follows:
\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

>>>>

To measure bias in word embeddings, @Caliskan2017semanticsBiases proposed the Word Embedding Association Test (WEAT). The idea is that the measure of  biases between two sets of target words, $X$ and $Y$, (we call them protected words) should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we'll call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words. WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. If the person's attitude towards given pair of concept is to be interpreted as neutral, there should be no noticeable task completion time difference, and the final value from the formula should be around 0.
Let $f$ be a similarity measure (usually, cosine similarity). The association difference for a term $t$ is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align}
\noindent then, the association difference between $A$ a $B$ is:
\begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align}
\noindent
$s(X,Y,A,B)$ is the statistic used in the signifcance test, and the $p$-value obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$. The effect size is computed by normalizing the difference in means as follows:
\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

### 2021-06-10 paperDraft/paperDraft.tex
first draft of abstract

To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
(2017) proposed the Word Embedding Association Test (WEAT). The idea is
that the measure of biases between two sets of target words, \(X\) and
\(Y\), (we call them protected words) should be quantified in terms of
the cosine similarity between the protected words and attribute words
coming from two sets of stereotype attribute words, \(A\) and \(B\)
(we'll call them attributes). For instance, \(X\) might be a set of male
names, \(Y\) a set of female names, \(A\) might contain stereotypically
male-related career words, and \(B\) stereotypically female-related
family words. WEAT is a modification of the Implicit Association Test
(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
almost the same word sets, allowing for a \emph{prima facie} sensible
comparison with bias in humans. Let \(f\) be a similarity measure
(usually, cosine similarity). The association difference for a term
\(t\) is:

@@

Caliskan et al. (2017) show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. Lauscher \& Glavas (2019) extended the methodology to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of similarty does not make much difference, while the
bias effects vary greatly accross embedding models (interestingly, with
social media-text trained embeddings being less biased than those based
on Wikipedia).

A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
(2018), who employ word embeddings trained od corpora from different
decades to study the shifts in various biases. For instance, to compute
the occupational embeddings bias for women the authors first compute the
average vector of vector emeddings of words that represent women (e.g.
``she'', ``female''), then calculate the Euclidean distance between this
mean vector and words for occupations. Then they take the mean of these
distances and subtract from it the analogously obtained mean for the
average vector of vector embeddings of words that represent men.
Formally they take the relative norm distance between \(X\) and \(Y\) to
be:

@@

\begin{itemize}
\item
  Seemingly high bias measures might arise even if the underlying
  processes actually have the same mean.
\item
  Even if the mean remains the same, non-negligible changes can result
  from a shift in the standard deviation of the original process, and
  the change might go in the opposite direction than a visualisation of
  datapoints might suggest, because with the decrease of standard
  deviation, the factor decreases and the bias increases.
\item
  Even if the underlying means are the same, but the variation is
  different, the bias metric in the long run could tend toward a
  different value:
\end{itemize}

>>>>

To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
(2017) proposed the Word Embedding Association Test (WEAT). The idea is
that the measure of biases between two sets of target words, \(X\) and
\(Y\), (we call them protected words) should be quantified in terms of
the cosine similarity between the protected words and attribute words
coming from two sets of stereotype attribute words, \(A\) and \(B\)
(we'll call them attributes). For instance, \(X\) might be a set of male
names, \(Y\) a set of female names, \(A\) might contain stereotypically
male-related career words, and \(B\) stereotypically female-related
family words. WEAT is a modification of the Implicit Association Test
(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
almost the same word sets, allowing for a \emph{prima facie} sensible
comparison with bias in humans. If the person's attitude towards given
pair of concept is to be interpreted as neutral, there should be no
noticeable task completion time difference, and the final value from the
formula should be around 0. Let \(f\) be a similarity measure (usually,
cosine similarity). The association difference for a term \(t\) is:

@@

Caliskan et al. (2017) show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. Lauscher \& Glavas (2019) extended the methodology to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of similarity does not make much difference, while the
bias effects vary greatly across embedding models (interestingly, with
social media-text trained embeddings being less biased than those based
on Wikipedia).

A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
(2018), who employ word embeddings trained on corpora from different
decades to study the shifts in various biases. For instance, to compute
the occupational embeddings bias for women the authors first compute the
average vector of vector emeddings of words that represent women (e.g.
``she'', ``female''), then calculate the Euclidean distance between this
mean vector and words for occupations. Then they take the mean of these
distances and subtract from it the analogously obtained mean for the
average vector of vector embeddings of words that represent men.
Formally they take the relative norm distance between \(X\) and \(Y\) to
be:

@@

\begin{itemize}
\item
  Seemingly high bias measures might arise even if the underlying
  processes actually have the same mean.
\item
  Even if the mean remains the same, non-negligible changes can result
  from a shift in the standard deviation of the original process, and
  the change might go in the opposite direction than a visualisation of
  datapoints might suggest, because with the decrease of standard
  deviation, the factor decreases and the bias increases.
\item
  Even if the underlying means are the same, but the variation is
  different, the bias metric in the long run could tend toward a
  different value:
\item
  The lack of control group in the paper and our analysis indicates that
  without neutral baseline it is difficult to interpret the
  effectiveness of the metric.
\end{itemize}

### 2021-06-10 thesis/_book/_main.tex
first draft of abstract

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(plyr)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{mapvalues}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass, }\DataTypeTok{from =} \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass), }\DataTypeTok{to =} \KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
\CommentTok{#levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

>>>>

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\CommentTok{#head(religion)}
\CommentTok{#library(plyr)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
\CommentTok{#levels(religion$wordClass)}
\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
\CommentTok{#                          from = levels(religion$wordClass),}
\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\chapter{Discussion}\label{discussion}
